---
title: "Spring_project"
author: "Marianna Baranovskaia"
date: "21 05 2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyr)
library(dplyr)
library(reshape2)
```

## Download the data
The data are downloaded, 2 data sets are joined with _rbind()_, and the structure is corrected
``` {r data}
kozak_dataset1 <-  read.csv("gnomad_hg38_af5pct_in_gencode_kozak_CDSstrands_snps_noborders_pos_sorted_annot_combo_F_clnsig_dataset.txt", sep = '\t', header = T)
kozak_dataset2 <- read.csv("clinvar_in_gencode_kozak_CDSstrands_snps_F_noborders_pos_sorted_annot_combo_clnsig_dataset.txt", sep = '\t', header = T)

kozak_dataset <- rbind(kozak_dataset1, kozak_dataset2)

kozak_dataset$variant_annotation = as.factor(kozak_dataset$variant_annotation)
kozak_dataset$Change_description = as.factor(kozak_dataset$Change_description)
kozak_dataset$Kozak_type = as.factor(kozak_dataset$Kozak_type)
kozak_dataset$Clin_Sig = as.factor(kozak_dataset$Clin_Sig)
kozak_dataset$Gene = as.factor(kozak_dataset$Gene)
kozak_dataset$Ref_Kozak_efficiency = as.numeric(kozak_dataset$Ref_Kozak_efficiency)
kozak_dataset$Ref_Kozak_lower = as.numeric(kozak_dataset$Ref_Kozak_lower)
kozak_dataset$Ref_Kozak_upper = as.numeric(kozak_dataset$Ref_Kozak_upper)
kozak_dataset$Alt_Kozak_efficiency = as.numeric(kozak_dataset$Alt_Kozak_efficiency)
kozak_dataset$Alt_Kozak_lower = as.numeric(kozak_dataset$Alt_Kozak_lower)
kozak_dataset$Alt_Kozak_upper = as.numeric(kozak_dataset$Alt_Kozak_upper)
kozak_dataset$Relative_efficiency = as.numeric(kozak_dataset$Relative_efficiency)

summary(kozak_dataset)

str(kozak_dataset)
```
The preprocessed data set is written in the file.
``` {r write01}
write.csv(kozak_dataset, "sum_dataset_01.csv")
```

##  Subset of the variants with known pathogenicity
I should analyse on this step of the project the variants with clear pathogenicity: Benign and Pathogenic or Likely Pathogenic or Pathogenic/Likely Pathogenic

``` {r data02}
known_sign = c('Pathogenic', 'Likely_pathogenic', 'Pathogenic/Likely_pathogenic', 'Benign')
kozak_dataset_short <-subset(kozak_dataset, Clin_Sig %in% known_sign)
kozak_dataset_short$pathogenic <- kozak_dataset_short$Clin_Sig %in% c('Pathogenic', 'Likely_pathogenic', 'Pathogenic/Likely_pathogenic')
head(kozak_dataset_short)

nrow(kozak_dataset_short[kozak_dataset_short$Clin_Sig == 'Benign', ]) 
nrow(kozak_dataset_short[kozak_dataset_short$Clin_Sig %in% c('Pathogenic', 'Likely_pathogenic', 'Pathogenic/Likely_pathogenic'), ])
nrow(kozak_dataset_short[kozak_dataset_short$Kozak_type == 'not_AUG_Kozak', ]) 
nrow(kozak_dataset_short[kozak_dataset_short$Kozak_type == 'AUG_Kozak', ]) 
nrow(kozak_dataset_short) 
```

This data set is written in the file too.
``` {r write02}
write.csv(kozak_dataset_short, "sum_dataset_known.csv")
```

## Vizualization for the big data set

Distributions of Kozak sequence efficiencies:

``` {r plots01, warning=FALSE}
plot_F_01 <- ggplot(kozak_dataset_short)+
  geom_histogram(aes(x=Ref_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_01

plot_F_02 <- ggplot(kozak_dataset_short)+
  geom_histogram(aes(x=Alt_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_02

plot_F_03 <- ggplot(kozak_dataset_short)+
  geom_histogram(aes(x=Relative_efficiency))+
  theme(legend.position="bottom", legend.box = "horizontal")
plot_F_03

plot_F_04 <- ggplot(kozak_dataset_short)+
  geom_histogram(aes(x=variant_annotation, fill = pathogenic), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal", 
        legend.text=element_text(size=14),
        axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14),
        axis.text.x=element_text(size=14, angle = 90),
        axis.text.y=element_text(size=14))+
  ylab("Count")+
  xlab("Variant annotation")
plot_F_04

plot_F_05 <- ggplot(kozak_dataset_short)+
  geom_histogram(aes(x=Kozak_type, fill = pathogenic), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")
plot_F_05
```
Saving the plots silently =)
``` {r plots_write01, include=FALSE}
jpeg(file="plot_F_01.jpeg")
plot_F_01
dev.off()

jpeg(file="plot_F_02.jpeg")
plot_F_02
dev.off()

jpeg(file="plot_F_03.jpeg")
plot_F_03
dev.off()

jpeg(file="plot_F_04.jpeg")
plot_F_04
dev.off()

jpeg(file="plot_F_05.jpeg")
plot_F_05
dev.off()
```

Significant part of the data belongs to the 'not_AUG_Kozak' subset but we work now only with AUG Kozak sequences.

## "AUG_Kozak" subset
``` {r data03}
kozak_dataset_short_AUG <- subset(kozak_dataset_short, Kozak_type == 'AUG_Kozak')
nrow(kozak_dataset_short_AUG) 
nrow(kozak_dataset_short_AUG[kozak_dataset_short_AUG$pathogenic == TRUE, ]) 
nrow(kozak_dataset_short_AUG[kozak_dataset_short_AUG$pathogenic == FALSE, ]) 
```

This data set is written in the file too.
``` {r write03}
write.csv(kozak_dataset_short_AUG, "sum_dataset_known_AUG.csv")
```

## Vizualization for the AUG_Kozak data set
``` {r plots02}

plot_F_06 <- ggplot(kozak_dataset_short_AUG)+
  geom_histogram(aes(x=Ref_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_06

plot_F_07 <- ggplot(kozak_dataset_short_AUG)+
  geom_histogram(aes(x=Alt_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_07

plot_F_08 <- ggplot(kozak_dataset_short_AUG)+
  geom_histogram(aes(x=Relative_efficiency))+
  theme(legend.position="bottom", legend.box = "horizontal")
plot_F_08

plot_F_09 <- ggplot(kozak_dataset_short_AUG)+
  geom_histogram(aes(x=variant_annotation, fill = pathogenic), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal", 
        legend.text=element_text(size=14),
        axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14),
        axis.text.x=element_text(size=14, angle = 90),
        axis.text.y=element_text(size=14))+
  ylab("Count")+
  xlab("Variant annotation")
plot_F_09

```

Saving the plots silently =)
``` {r plots_write02, include=FALSE}
jpeg(file="plot_F_06.jpeg")
plot_F_06
dev.off()

jpeg(file="plot_F_07.jpeg")
plot_F_07
dev.off()

jpeg(file="plot_F_08.jpeg")
plot_F_08
dev.off()

jpeg(file="plot_F_09.jpeg")
plot_F_09
dev.off()
```

## "Upstream+synonymous" subset
We decided to do this subset because 'no_start', 'missense' and 'nonsense' variants change the protein primary structure and can be pathogenic just because of this effect and not because of up/downregulation with Kozak sequence. 

``` {r data04}
locations = c('upstream', 'synonymous')
kozak_dataset_short_AUG_2 <- subset(kozak_dataset_short_AUG, variant_annotation %in% locations)
nrow(kozak_dataset_short_AUG_2)
nrow(kozak_dataset_short_AUG_2[kozak_dataset_short_AUG_2$pathogenic == TRUE, ]) 
nrow(kozak_dataset_short_AUG_2[kozak_dataset_short_AUG_2$pathogenic == FALSE, ])
```

## Vizualization for "Upstream+synonymous" subset
``` {r plots03}

plot_F_10 <- ggplot(kozak_dataset_short_AUG_2)+
  geom_histogram(aes(x=Ref_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_10

plot_F_11 <- ggplot(kozak_dataset_short_AUG_2)+
  geom_histogram(aes(x=Alt_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_11

plot_F_12 <- ggplot(kozak_dataset_short_AUG_2)+
  geom_histogram(aes(x=Relative_efficiency))+
  theme(legend.position="bottom", legend.box = "horizontal")
plot_F_12

plot_F_13 <- ggplot(kozak_dataset_short_AUG_2)+
  geom_histogram(aes(x=variant_annotation, fill = pathogenic), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal", 
        legend.text=element_text(size=14),
        axis.title.x=element_text(size=14),
        axis.title.y=element_text(size=14),
        axis.text.x=element_text(size=14, angle = 90),
        axis.text.y=element_text(size=14))+
  ylab("Count")+
  xlab("Variant annotation")
plot_F_13
```

Saving the plots silently =)
``` {r plots_write03, include=FALSE}
jpeg(file="plot_F_10.jpeg")
plot_F_10
dev.off()

jpeg(file="plot_F_11.jpeg")
plot_F_11
dev.off()

jpeg(file="plot_F_12.jpeg")
plot_F_12
dev.off()

jpeg(file="plot_F_13.jpeg")
plot_F_13
dev.off()
```
Which variants have the effect more than 50%?
``` {r data05}
subset(kozak_dataset_short_AUG_2, Relative_efficiency>1.5)

subset(kozak_dataset_short_AUG_2, Relative_efficiency<0.5)
```
## Vizualization for "Upstream+synonymous" subset (melted)
``` {r data06}
kozak_dataset_short_AUG_2_part_melted <- melt(kozak_dataset_short_AUG_2[, c('ID',  'pathogenic', 'Ref_Kozak_efficiency', "Alt_Kozak_efficiency")], id.vars=c("ID", 'pathogenic'))

kozak_dataset_short_AUG_2_part_melted2 <- melt(kozak_dataset_short_AUG_2[, c('ID',  'pathogenic', 'Kozak_variant_position', 'Ref_Kozak_efficiency', "Alt_Kozak_efficiency")], id.vars=c("ID", 'pathogenic', 'Kozak_variant_position'))
kozak_dataset_short_AUG_2_part_melted2$Kozak_variant_position <- as.factor(kozak_dataset_short_AUG_2_part_melted2$Kozak_variant_position)
```

``` {r plots04}

plot_F_14 <- ggplot(kozak_dataset_short_AUG_2_part_melted, aes(x=pathogenic, fill=variable))+
  geom_boxplot(aes(y=value))+
  theme(legend.position="bottom", legend.box = "horizontal")+
  ylab('Kozak sequence efficiency')+
  xlab('Pathogenic')
plot_F_14

plot_F_15 <- ggplot(kozak_dataset_short_AUG_2_part_melted2, aes(x=Kozak_variant_position, fill=variable))+
  geom_boxplot(aes(y=value))+
  theme(legend.position="bottom", legend.box = "horizontal")+
  ylab('Kozak sequence efficiency')+
  xlab('Kozak_variant_position')+
  facet_grid(rows = vars(pathogenic))
plot_F_15

plot_F_16 <- ggplot(kozak_dataset_short_AUG_2_part_melted2, aes(x=pathogenic, fill=variable))+
  geom_boxplot(aes(y=value))+
  theme(legend.position="bottom", legend.box = "horizontal")+
  ylab('Kozak sequence efficiency')+
  xlab('Pathogenic or not')+
  facet_grid(cols = vars(Kozak_variant_position))
plot_F_16
```

Saving the plots silently =)
``` {r plots_write04, include=FALSE}
jpeg(file="plot_F_14.jpeg")
plot_F_14
dev.off()

jpeg(file="plot_F_15.jpeg")
plot_F_15
dev.off()

jpeg(file="plot_F_16.jpeg")
plot_F_16
dev.off()
```

There seems to be no significant differences in the distributions here. 

## "Only significant" subset
The last subset is the variants which have non-intersected confidence intervals (i.e. with significant change in the translation efficiency)
``` {r data07}
kozak_dataset_short_AUG_3 <- subset(kozak_dataset_short_AUG_2, Relative_efficiency != 'NA')

kozak_dataset_short_AUG_3$is_significant <- c(NA * nrow(kozak_dataset_short_AUG_3))

kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency > 1, ]$is_significant <- 
  ifelse(kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency > 1, ]$Alt_Kozak_lower > 
           kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency > 1, ]$Ref_Kozak_upper, TRUE, FALSE)

kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency <= 1, ]$is_significant <- 
  ifelse(kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency <= 1, ]$Alt_Kozak_upper < 
           kozak_dataset_short_AUG_3[kozak_dataset_short_AUG_3$Relative_efficiency <= 1, ]$Ref_Kozak_lower, TRUE, FALSE)

sign_diff_Kozaks_AUG <- subset(kozak_dataset_short_AUG_3, is_significant == TRUE)
nrow(kozak_dataset_short_AUG_3)
nrow(sign_diff_Kozaks_AUG) 
nrow(sign_diff_Kozaks_AUG[sign_diff_Kozaks_AUG$pathogenic == TRUE,])
nrow(sign_diff_Kozaks_AUG[sign_diff_Kozaks_AUG$pathogenic == FALSE,])
```
## Vizualization of "Only significant" subset
``` {r plots05}

plot_F_16 <- ggplot(sign_diff_Kozaks_AUG)+
  geom_histogram(aes(x=Ref_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_16

plot_F_17 <- ggplot(sign_diff_Kozaks_AUG)+
  geom_histogram(aes(x=Alt_Kozak_efficiency), stat="count")+
  theme(legend.position="bottom", legend.box = "horizontal")+
  xlim(c(0,150))
plot_F_17

plot_F_18 <- ggplot(sign_diff_Kozaks_AUG)+
  geom_histogram(aes(x=Relative_efficiency))+
  theme(legend.position="bottom", legend.box = "horizontal")
plot_F_18
```

Saving the plots silently =)
``` {r plots_write05, include=FALSE}
jpeg(file="plot_F_16.jpeg")
plot_F_16
dev.off()

jpeg(file="plot_F_17.jpeg")
plot_F_17
dev.off()

jpeg(file="plot_F_18.jpeg")
plot_F_18
dev.off()
```

## Vizualization of "Only significant" subset (melted)
``` {r data08}
sign_diff_Kozaks_AUG_part_melted <- melt(sign_diff_Kozaks_AUG[, c('ID',  'pathogenic', 'Ref_Kozak_efficiency', "Alt_Kozak_efficiency")], id.vars=c("ID", 'pathogenic'))

sign_diff_Kozaks_AUG_part_melted2 <- melt(sign_diff_Kozaks_AUG[, c('ID',  'pathogenic', 'Kozak_variant_position', 'Ref_Kozak_efficiency', "Alt_Kozak_efficiency")], id.vars=c("ID", 'Kozak_variant_position', 'pathogenic'))
sign_diff_Kozaks_AUG_part_melted2$Kozak_variant_position <- as.factor(sign_diff_Kozaks_AUG_part_melted2$Kozak_variant_position)

```

``` {r plots06}

plot_F_19 <- ggplot(sign_diff_Kozaks_AUG_part_melted, aes(x=pathogenic, fill=variable))+
  geom_violin(aes(y=value))+
  geom_boxplot(aes(y=value), alpha=0.2)+
  theme(legend.position="bottom", legend.box = "horizontal", 
        legend.text = element_text(size=14),
        axis.title.x=element_text(size=14), 
        axis.title.y=element_text(size=14),
        axis.text.x=element_text(size=14), 
        axis.text.y=element_text(size=14))+
  ylab('Kozak sequence efficiency')+
  xlab('Pathogenic')+
  scale_fill_manual(values = c("cyan", "gray"))
plot_F_19

plot_F_20 <- ggplot(sign_diff_Kozaks_AUG_part_melted2, aes(x=pathogenic, y=value))+
  geom_violin(aes(fill=variable))+
  theme(legend.position="bottom", legend.box = "horizontal")+
  ylab('Kozak sequence efficiency')+
  xlab('Pathogenic or not')+
  facet_grid(cols = vars(Kozak_variant_position))+
  scale_fill_manual(values = c("cyan", "gray"))+
  theme(legend.position="bottom", legend.box = "horizontal", 
        legend.text = element_text(size=14),
        axis.title.x=element_text(size=14), 
        axis.title.y=element_text(size=14),
        axis.text.y=element_text(size=14))+
  ylab('Kozak sequence efficiency')+
  xlab('Pathogenic')
plot_F_20
```



Saving the plots silently =)
``` {r plots_write06, include=FALSE}
jpeg(file="plot_F_19.jpeg")
plot_F_19
dev.off()

jpeg(file="plot_F_20.jpeg")
plot_F_20
dev.off()
```